Transfer learning for echelle spectroscopy

### 2021-11-16 paper/paper1/ms.tex
Add values to placeholders, prefactor discussion

We initialize the cloned model with a series of preprocessing steps.  We divide the entire spectrum by a black body of the same effective temperature $T_{\mathrm{eff}}$ as the model template.  The resulting signal typically possesses a smooth continuum variation.  An additional continuum flattening step ensures that subsequent spectral line finding steps get applied uniformly.  Here, we identify the continuum regions as the most prominent peaks in the specturm separated by a pixel distance of at least $15000-90000$ pixels from adjacent peaks, yielding between 60 and 5 points.  We then fit an $n_{\mathrm{poly}} = 3^{rd}-5^{th}$ order polynomial to these peaks, evaluate the polynomial on the entire bandwidth, and divide the spectrum by this trend.  Visual inspection of this continuum-flattened spectrum typically possesses variations at the $<0.2\%$ level.  Tuning the pixel distance and polynomial order can lower the undulations to $<0.05\%$ level for some input spectra.  The high-pass filtering step should be set to capture the genuine spectral shape, without over-fitting broad line-wings such as those in deep Hydrogen and sodium lines.  We have experimentally found it helpful to hand-tune models of exceptionally broad and non-standard line-wings with an optional pre-processing step, not described here.

We emphasize that a recreation of the unvarnished input spectrum can be obtained by multiplying the continuum-flattened signal by the ``perturbed black body'', $\mathsf{B}(\bm{\lambda}_s)$, that symbolizes the black body modulated by continuum opacity or broad-band radiative transfer effects.  This smooth spectrum may be useful for applications that need to keep track of broad-band flux, such as low resolution spectrographs, or regions with molecular bandheads.  The ``perturbed black body'' continuum model $\mathsf{B}$ contains  $n_{\mathrm{poly}}+1$ fixed lookup parameters, plus the input $T_{\mathrm{eff}}$.

@@

For this paper, we illustrate examples for two \'echelle spectrographs with particularly large spectral grasp: the Habitable Zone Planet Finder \citep[HPF,][]{2014SPIE.9147E..1GM} on the Hobby-Eberly Telescope at McDonald Observatory in Fort Davis, Texas; and the Immersion Grating Infrared Spectrograph \citep[IGRINS,][]{park14}
currently on the Gemini South Telescope on Cerro Pach\'on in Chile.  The $R=55,000$ HPF has a native bandwidth of $x-y$, that we expand to $x-y^*$ including the edge buffers.  IGRINS has two cameras for $H$ and $K$ band, with the combined spectrum spanning $z-y$ including the edge buffers and the region in-between the two cameras, all at a resolving power of $(R=45,000)$.  The spectrograph acquisition, reduction, and post-processing steps yield data $\mathsf{D}(\bm{\lambda}_{d})$, where $\bm{\lambda}_{d}$ is the wavelength vector at the instrumental resolution and sampling of each instrument, generally much coarser than the resolution and sampling grid of the precomputed synthetic spectra.  The data wavelength vector may also contains gaps between \'echelle orders, whereas the precomputed grid is usually contiguous.  HPF may have up to $2048\times28=57,344$ pixels, and IGRINS has typically about $75,000$ pixels, after common truncation of noisy edge pixels.

@@

Figure \ref{fig_Nlines_vs_teff} shows how the number of detected lines $N_{\mathrm{lines}}$ scales with effective temperature and prominence threshold $P_{rom}$ for the \texttt{PHOENIX} grid, truncated to the bandwidths-plus-buffers for HPF and IGRINS.  The HPF-truncated spectra have $N_s=XX$ and the IGRINS-truncated spectra have $N_s=XX$.

@@

\begin{eqnarray}
  \mathsf{M}(\bm{\lambda}_s) = {\displaystyle \prod_{j=1}^{N_{\mathrm{lines}}}  (1-a_j \mathsf{V}(\bm{\lambda}_s-\lambda_{c,j}, \sigma_j, \gamma_j) )} \label{equation1}
\end{eqnarray}

where $\mathsf{V}(\lambda, \sigma_j, \gamma_j)$ is the normalized Voigt profile with Gaussian standard deviation $\sigma$, Lorentzian half-width $\gamma$, at line center position $\lambda_c$, for the $j^{th}$ spectral line.  The amplitude $a$ is always expected to be positive for absorption lines, and is equal to the area under the curve of the line since the Voigt profile is normalized.

@@

As seen in Figure \ref{fig_Nlines_vs_teff}, the number of lines can exceed $XX$, meaning the model has over $XX \times 3 = ZZ$ free parameters.  Fitting that large number of parameters is difficult with conventional optimizers, which struggle to converge after hundreds of parameters\authorcomment1{Citation to limits of optimizers}.

>>>>

We initialize the cloned model with a series of preprocessing steps.  We divide the entire spectrum by a black body of the same effective temperature $T_{\mathrm{eff}}$ as the model template.  The resulting signal typically possesses a smooth continuum variation.  An additional continuum flattening step ensures that subsequent spectral line finding steps get applied uniformly.  Here, we identify the continuum regions by the most prominent peaks in the specturm separated by a distance of at least 5,000$-$90,000 pixels from adjacent peaks, yielding between 60 and 5 control points.  We then fit an $n_{\mathrm{poly}} = 3^{rd}-15^{th}$ order polynomial to these peaks, evaluate the polynomial on the entire bandwidth $\mathsf{P}(\bm{\lambda}_s)$, and divide the spectrum by this trend.  Visual inspection of this continuum-flattened spectrum typically possesses variations at the $<1\%$ level.  Tuning the pixel distance and polynomial order can lower the undulations to $<0.05\%$ level for some input spectra.  We found a pixel distance of 50,000 and $3^{rd}$ order polynomial worked across the range of effective temperatures for HPF's bandwidth. The IGRINS spectra had more significant continuum opacity and bandheads to accomodate and therefore needed to capture finer scale variation: 5,000 pixels and a $9^{th}$ order polynomial appeared adequate over most of the effective temperature range.  This high-pass filtering step should be set to capture the genuine spectral shape, without over-fitting broad line-wings such as those in deep Hydrogen and sodium lines.  We have experimentally found it helpful to hand-tune models of exceptionally broad and non-standard line-wings with an optional pre-processing step, not described here.

We emphasize that a recreation of the unvarnished input spectrum can be obtained by multiplying the continuum-flattened signal by the ``perturbed black body'', $\mathsf{B}(\bm{\lambda}_s)\cdot \mathsf{P}(\bm{\lambda}_s)$, that symbolizes the black body modulated by continuum opacity or broad-band radiative transfer effects.  This smooth spectrum may be useful for applications that need to keep track of broad-band flux, such as low resolution spectra, or regions with molecular bandheads.  The ``perturbed black body'' continuum model contains $n_{\mathrm{poly}}+1$ fixed-but-possibly tunable lookup parameters, plus the fixed input $T_{\mathrm{eff}}$.

@@

For this paper, we illustrate examples for two \'echelle spectrographs with particularly large spectral grasp: the Habitable Zone Planet Finder \citep[HPF,][]{2014SPIE.9147E..1GM} on the Hobby-Eberly Telescope at McDonald Observatory in Fort Davis, Texas; and the Immersion Grating Infrared Spectrograph \citep[IGRINS,][]{park14}
currently on the Gemini South Telescope on Cerro Pach\'on in Chile.  The $R=55,000$ HPF has a native bandwidth of $8079-12785$~\AA, that we expand to $8049-12815$~\AA~including the edge buffers.  IGRINS has two cameras for $H$ and $K$ band, with the combined spectrum spanning $14267-25217\;$\AA~ including the edge buffers and the region in-between the two cameras, all at a resolving power of $(R=45,000)$.  The spectrograph acquisition, reduction, and post-processing steps yield data $\mathsf{D}(\bm{\lambda}_{d})$, where $\bm{\lambda}_{d}$ is the wavelength vector at the instrumental resolution and sampling of each instrument, generally much coarser than the resolution and sampling grid of the precomputed synthetic spectra.  The data wavelength vector may also contain gaps between \'echelle orders, whereas the precomputed wavelength coordinates are usually contiguous.  HPF may have up to $2048\times28=$57,344 pixels, and IGRINS has typically about 75,000 pixels, after common trimming of noisy edge pixels and unusable telluric regions.

@@

Figure \ref{fig_Nlines_vs_teff} shows how the number of detected lines $N_{\mathrm{lines}}$ scales with effective temperature and prominence threshold $P_{rom}$ for the \texttt{PHOENIX} grid, truncated to the bandwidths-plus-buffers for HPF and IGRINS.  The HPF-truncated spectra have $N_s=$335,849 and the IGRINS-truncated spectra have $N_s=$330,052.

@@

\begin{eqnarray}
  \mathsf{M}(\bm{\lambda}_s) = {\displaystyle \mathsf{P} \prod_{j=1}^{N_{\mathrm{lines}}}  (1-a_j \mathsf{V}(\bm{\lambda}_s-\lambda_{c,j}, \sigma_j, \gamma_j) )} \label{equation1}
\end{eqnarray}

where $\mathsf{V}(\lambda, \sigma_j, \gamma_j)$ is the normalized Voigt profile with Gaussian standard deviation $\sigma$, Lorentzian half-width $\gamma$, at line center position $\lambda_c$, for the $j^{th}$ spectral line.  The amplitude $a$ is always expected to be positive for absorption lines, and is equal to the area under the curve of the line since the Voigt profile is normalized.  The optional Prefactor term $\mathsf{P}$ is fixed at $1$ by default, but can be allowed to vary to fine-adjust imperfections left over from the polynomial-based continuum flattening procedure.

@@

As seen in Figure \ref{fig_Nlines_vs_teff}, the number of lines can exceed 7,000, meaning the model has over 7,000 $\times 3 =$ 21,000 free parameters.  Fitting that large number of parameters is difficult with conventional optimizers, which struggle to converge after hundreds of parameters\authorcomment1{Citation to limits of optimizers}.
